Differences in Homicide Case Information Indicates Why Justice is Not Served

### 2024-12-15 paper/paper.qmd
Finished making changes to paper

The dataset we used for this paper comes from The Washington's Post's GitHub repository, "How The Post mapped unsolved murders", which is also known as the "Unsolved Homicide Database" [@datahomicides]. We used the statistical programming language R [@citeR], tidyverse [@citetidyverse], janitor [@citejanitor], lubridate [@citelubridate], dplyr [@citedplyr], ggplot2 [@citeggplot2], arrow [@citearrow], testthat [@citetestthat], and knitr [@citeknitr] to retrieve, clean, test, and analyze the dataset. To construct, test, and analyze our model, we used the following packages: rstanarm [@citerstanarm], modelsummary [@citemodelsummary], and car [@citecar]. The causal model diagram created to understand the relationship between predictor variables, the outcome variable, and a confounder used DiagrammeR [@citeDiagrammeR], rsvg [@citersvg], DiagrammeRsvg [@citeDiagrammeRsvg] and png [@citepng].

The Washington Post's Unsolved Homicide Database is a dataset compiled by Washington Post reporters that contains over 52000 homicides in the United States (US) from 50 of the largest US cities from 2007 to 2017 [@datahomicides]. This dataset includes information about the victim such as their name, sex, race, and age as well as geographic and temporal information of the homicide. The Washington Post was interested in using the information compiled to map unsolved homicides across the United States in major cities from 2007 to 2017 [@citewashingtonpostmap]. Another dataset we had considered using was another one compiled by The Washington Post on school shootings across the US and approaching our problem with a different perspective on the characteristics of the perpetrator. However, due to there only being 416 observations and numerous observations missing information, we forgo using the dataset.

The raw dataset we retrieved from The Washington Post using a script that downloads the CSV file from their GitHub repository contains 52,179 observations with each observation being a homicide case. However, since we narrowed our scope to focus on the two of the most populated US cities, Los Angeles and Chicago, the number of observations in our cleaned dataset ended up being 6,307. Our data look as follows:

@@

@tbl-dataset-preview and @tbl-dataset-statistics indicates our variables of interest, which are the following: victim_race, victim_age, victim_sex, city, disposition, year, month, and arrest_was_not_made. victim_race represents the race of the homicide victim, which can be "White" (376 observations), "Hispanic" (1,763 observations), "Black" (4,063 observations), "Asian" (43 observations), and "Other" (62 observations). victim_age signifies the age of the homicide victim at the time of their death and is defined as a whole number. @tbl-dataset-statistics shows that the mean victim_sex is the sex of the homicide victim where they are either identified as a "female" or "male". The city variable defines the city the victim is reported to have been found in. The disposition variable is the specific status of a homicide case where a case can fall in either of the following three status: "Closed by arrest", "Open/No arrest", and "Closed without arrest". The year variable represents a year from 2010 to 2017 that indicates the year the homicide took place. Following this, the month variable represents the month a homicide took place. arrest_was_not_made is a variable that was constructed based on the disposition variable with "Closed by arrest" being converted to a 0 and "Open/No arrest" and "Closed without arrest" being converted to a 1. arrest_was_not_made indicates the status of a homicide case as either being unsolved, which is denoted by a 1, and solved, which is denoted by a 0.

However, our dataset has limitations. There was only data available from 2010 onwards for Los Angeles provided by The Washington Post and so that limited the number of homicides we could look at for both Chicago and Los Angeles. Also, not all victims were able to be identified in some homicide cases and unknown attributes of the victim such as their age, sex, and gender were indicated with the text "Unknown" in the dataset. However, this causes issues with data type compatibility such as the value "Unknown" being a character type being under the victim_age column, which has a data type of integer. This would lead to issues with our model providing accurate estimates. We decided to remove cases during our data cleaning where victim's demographic information is missing at least one of the three columns, victim_age, victim_sex, victim_race.

@@

The Federal Bureau of Investigation (FBI) has a program called the Uniform Crime Reporting (UCR) Program to generate statistics for the public [@citeFBIinfo]. The FBI originally had a system under the UCR called Summary Reporting System (SRS), which obtained details about different crimes taking place from law enforcement agencies nationwide such as victim information [@citeFBIinfo]. However, the SRS was replaced with a new system called the National Incident-Based Reporting System (NIBR) in 2021 that obtained more details about various crimes [@citeFBIinfo]. For The Los Angeles Police Department, after a homicide case occurs, the investigating team handwrites information into physical crime reports with details like the type of crime, the premise the crime occured at, and the age and ethnicity of the victim [@citeLApolice]. The crime reports are then transcribed into a digital format and then sent to the SRS monthly [@citeLApolice]. After a homicide occurs in Chicago, The Chicago Police Department uses a system called the Chicago Police Department's CLEAR (Citizen Law Enforcement Analysis and Reporting) system to report on details such as the victim, if an arrest was made or not, and the location the crime took [@citeChicagoPolice]. This information is then reported to the FBI monthly under the UCR program [@citeChicagoPolice].

Reporters from the Washington Post then obtained the data from the FBI specifically on homicides from 50 largest US cities from 2007 to 2017, which can be access through the UCR publications page on the FBI website [@datahomicides]. They selected the 50 largest cities based the size of the city in 2012 [@datahomicides]. They also obtain data about homicide counts and closure rates through papers and compare these values with the ones from the FBI dataset for accuracy [@datahomicides]. The Washington Post would also use public records like medical examiner reports, death certificates, and court records, to fill in any missing information since some departments only report partial information to the FBI [@datahomicides]. The Washington Post defined cases to be closed without arrest when they are reported by the police as "exceptionally cleared" where there is evidence of who the perpetrator is but an arrest is not possible because of reason like they has died [@datahomicides]. They also define cases to be closed by arrest if the police reported it to be and other cases are defined to be open/no arrest [@datahomicides].

## Outcome variable
The outcome variable we are interested in looking at with our model and our analysis is the arrest_was_not_made variable. We use this variable to compare homicide case characteristics of solved and unsolved homicides.

@@

# Model {#sec-model}
The model we implemented was a Bayesian logistic regression model. This model was constructed after we saw patterns between homicide case information and a homicide case going unsolved in our analysis. We are interested in seeing if certain characteristics of a homicide case such as the victim's perceived characteristics (sex, gender, age) and the year and city the victim is found impacts the likelihood of their case going unsolved.

@@

We define $y_i$ to be the status of the homicide case where 1 means the homicide is unsolved (case is still open / been closed without arrest) and 0 means the homicide is solved (case has been closed with arrest). $\pi_i$ represents the probability of the homicide being solved. $\mbox{logit}(\pi_i)$ indicates the log-odds of the homicide being unsolved. Now looking at the coefficients $b_i$ and predictor variables, $\beta_0$ is the intercept of our model and is the log-odds when all predictor variables are equal to 0. $\mbox{victim\_race}_i$ signifies the race of the victim, which could be either "Asian", "Black", "Hispanic", "White", and "Other". In our model, we use "White" as the baseline for $\mbox{victim\_race}_i$ to see if being part of a minority impacts if the case goes unsolved or not. $\beta_1$ is the coefficient that represents the log-odds when $\mbox{victim\_race}_i$ changes. $\mbox{victim\_age}_i$ is the victim's age, which is a whole number. $\beta_2$ is the log-odds when $\mbox{victim\_age}_i$ increases by 1 year. $\mbox{victim\_sex}_i$ represents the sex of the victim, where in the dataset it is either "female" or "male" and $\beta_3$ is the log-odds when $\mbox{victim\_sex}_i$ changes. $\mbox{city}_i$ is the city the victim was reported to be killed in, which from our dataset could be either "Chicago" or "Los Angeles". $\beta_4$ is the coefficient that stands for the log-odds when $\mbox{city}_i$ changes. We define $\mbox{year}_i$ to be the year the homicide was reported to have occured from 2010 to 2017. $\beta_5$ indicates the log-odds when $\mbox{year}_i$ increases by 1 year.

Our model runs in R [@citeR] using the rstanarm package [@citerstanarm]. For our model's priors, we use the default priors provided by rstanarm [@citerstanarm]. Diagnostics for the model such as in posterior predictive check, trace and Rhat plots can be found in @sec-model-details.

@@

In our model, we assumed there is a relationship between the outcome variable, homicide is unsolved, with homicide case information like demographic (sex, race, and age of victim), geographic (city), and temporal (year) information, which are the predictor variables. For the demographic data, we decided to keep the grouping provided by The Washington Post such as for sex it was "female" and "male" and race it was "White", "Black", "Hispanic", "Asian", and "Other. However, we did remove victims who has any demographic information that falls under the "Unknown" grouping from our dataset and did not run our regression model on these observations. The reason for this is due to factors such as the data type of the predictor and keeping consistency across all observation and removing any unknown values. For example, our predictor variable, victim's age is a integer data type but it contained the string "Unknown" in the raw dataset. So the values "Unknown" is removed in our final dataset and not used to train our model. Since our outcome variable is binary for our model, we constructed the arrest_was_not_made column for it based on the disposition variable to reflect it and considered dispositions with values like "Open/No arrest" and "Closed without arrest" to be 1 and "Closed by arrest" to be 0.

We used a logistic regression model in a Bayesian framework due to the fact that our outcome is binary and predicts if a homicide is unsolved or not. Another model we considered is a logistic regression model with an instrumental variable. Introducing a instrumental variable into our model could have potentially provided a more accurate model and given us more consistent coefficient estimates as noted by @citeucdavis. However with the available information we had about each case, there was no candidate instrumental variable that impacted at least one variable in our data and not influence the outcome of the case being solved or unsolved. Thus, the model would fail the "Exclusion Restriction" assumption mentioned by @citetellingstories. We also went through different pairs and groupings of variables and how there was not a strong relationship between variables that is relevant and consequently fail @citetellingstories's "Relevance" assumption. We also have known treatment variable/predictor variables that can be used to measure the outcome variable and therefore, the instrumental variable is less likely to be necessary [@citetellingstories].

Performing root mean square error (RSME) calculations on our model in @tbl-model-result yields a RSME value of about 0.45. RSME represents the difference between the model's predicted values and observed values with 0 meaning that the predicted and observed values are the same [@citejim]. Since the model's RSME value is close to 0, we can say our model is able to predict values with less error compared to other models that have a higher RSME value.

@@

However, our model has limitations and there are situations where this model would not work. @fig-causal-model shows that there is a confounder, "resources of investigating team" between the predictor variables, city and year and the outcome variable, case being unsolved. We define "resources of the investigating team" to include any of the following: the amount of people on the team investigating the case, time available allotted to investigate the case, amount of open cases for the team, cost, skill and education levels of members, etc. We currently do not have any information available about the resources of the investigating team for the case and further investigation is needed. We proposed an idealized survey we would conduct to collect the necessary data to further understand the relationship between homicide case characteristics and unsolved homicides in @sec-ideal-survey. If we have information available about the resources of the investigating team for the case, the current model would not work and would need to be revised. This is due to the interaction between the city, year, and outcome variables with the confounder.

@@

@fig-disposition-bar and @tbl-disposition-table shows that the number of homicide cases that are closed by arrest at 1,109 cases is close to the number of homicide cases that are open at 1,087 cases from 2010 to 2017. On the hand, there is a larger gap for Chicago between the number of homicides that are closed by arrest at 947 cases with the ones that are closed without arrest and open/no arrest at 216 and 2,948 cases, respectively.

@@

@fig-victim-age-histogram shows that the distribution of age of homicide victims is relatively uniform. It appears that the victims of unsolved and solved homicides are about the same age if one were not to consider ages above 75. However, if we were to consider age 75 onwards, it appears that the distribution is a bit right-skewed. We can see that from @tbl-victim-age-statistics, the youngest homicide victim is 1 years old and the oldest is 94 years old. The median age of homicide victims is 44 years old with the mean age being 45 years old.

@@

```{r}
#| label: fig-model-result-ci
#| fig-cap: The credible intervals (line) for coefficient estimates (dot) of predictor variables for homicides that go unsolved from 2010 to 2017.
#| echo: false
#| warning: false
#| message: false
# Specify credible interval to be 90% and present model results
modelplot(unsolved_model_results, conf_level = 0.9) +
  labs(x = "90 per cent credible interval")
```
@tbl-model-result and @fig-model-result-ci presents the results from our logistic regression model in the Bayesian framework and the 90% credible intervals of each predictor, respectively. From @tbl-model-result, the intercept $\beta_0$ of -165.865 indicates that when the homicide victim is White, since "White" is the baseline for the victim_race predictor, and victim_age is 0 and year being 2010 with its coefficient being 0 and the victim's sex is female, the homicide is more likely to be solved. @tbl-model-result indicates that the coefficient estimate of victim_race $\beta_1$ is 0.168 when the victim is Asian. This means that the log-odds of a homicide case being unsolved increases by 0.168 when the victim is Asian while other predictor variables stays fixed. Following this, when the victim's race is Black or Hispanic, the log-odds of a homicide being unsolved increases by 0.567 and 0.434, respectively. This indicates that the likelihood of a homicide going unsolved increases if the victim's race is Asian, Black, or Hispanic relative to a victim's race being White. On the other hand, when the victim's race falls under "Other", the log-odds of a homicide case being unsolved decreases by 0.296. This indicates that the likelihood of a homicide being unsolved decreases if the victim's race falls under "Other" relative to a victim's race being White. However, since the 90% credible interval for the coefficient estimate of victim_race appears to be close 0 as seen in @fig-model-result-ci, this means that a victim's race has a weak likelihood of impacting the outcome of their case being unsolved.

@tbl-model-result shows that since the coefficient estimate of victim_age $\beta_2$ is -0.006, this means the log-odds of a homicide case being unsolved decreases by 0.006 when the age of the victim increases by 1 year as other predictor variables stay constant. This means victims of unsolved homicides are likely on the more younger side. @fig-model-result-ci also indicates that $\beta_2$'s 90% credible interval is close to 0 or includes 0 in its interval, indicating there is a chance the victim's age does not have an influence the outcome of a homicide case being unsolved. With the coefficient estimate of victim_sex $\beta_3$ being 0.662 in @tbl-model-result, it indicates that the log-odds of a homicide case being unsolved increases by 0.662 when the victim is male while other predictors are fixed. This suggests that the likelihood of the a homicide case being unsolved increases when the victim is a male. Looking at @fig-model-result-ci, the credible interval for $\beta_3$ is also slightly above 0. @tbl-model-result also shows that the log-odds of a homicide being unsolved decreases by 1.089 as indicated by the coefficient estimate $\beta_4$ when city is set to Los Angeles with other predictors being fixed. This indicates that more unsolved homicide cases are likely to occur in Chicago instead. @fig-model-result-ci also shows that $\beta_4$'s credible interval is below 0. Looking at the year predictor and its coefficient estimate, the log-odds of a homicide case being unsolved increases by 0.083 as noted by the coefficient estimate $\beta_5$ when the year increases by 1 year while the other predictors stay constant. This indicates that cases are more likely to be unsolved in the later years between 2010 to 2017. @fig-model-result-ci also shows that $\beta_5$'s credible interval is slightly above 0 or includes 0 indicating that the year has a weak likelihood of indicating if a homicide is unsolved.

@@

In @sec-results, we analyzed homicide case information like the victim's age, sex, and race and the year and city the homicide case occurred to see if there were any differences between homicide case information between unsolved and solved homicides from the year 2010 to 2017. We also built a Bayesian logistic regression model in @sec-model-results, to see the likelihood of the homicide case information impacting the homicide case outcome of going unsolved.

## Chicago likely to have more unsolved homicides than Los Angeles with a surge in unsolved cases from 2016 onwards

From @fig-city-bar and @tbl-city-table, it indicates that while Los Angeles have slightly more homicide cases that are solved than unsolved, 77% of Chicago's homicide cases are unsolved from 2010 to 2017. This is also seen in @tbl-model-result, where our model indicates that that unsolved homicides in Chicago are likely to go unsolved compared to Los Angeles. This implies potential issues with how the Chicago Police Department is managing homicide case closure. From @fig-month-bar and @fig-year-bar, we saw that most unsolved homicide cases occur in the middle and later half of the year and that there was jump in unsolved homicides in 2016 and 2017 compared to past years. @citeCNNChicago reported on January 2, 2017 that 2016 was the year with the most number homicides in the past 19 years due to gun violence in the city. This suggest that most of the unsolved cases could have occurred due to police departments like the Chicago Police Department being unable to keep up with the surge of homicide cases.

@@

@fig-victim-age-histogram indicates that the age of the homicide victim likely does not impact the outcome of their case be unsolved due to the uniformity of the distribution of ages. This is further seen in @fig-model-result-ci, where there is a chance the victim's age does not influence the outcome of a homicide case in our model. This implies that the age of the homicide victim does not impact if their case is unsolved or not. However, @fig-model-result-ci and @fig-victim-sex-bar indicates that female victims had a higher proportion of their homicide cases being solved while on the otherhand, male victims had a higher proportion of their homicide cases going unsolved. This suggest that males are more likely to have their case go unsolved.

## Disportionate number of homicide victims in unsolved cases in Chicago and Los Angeles from 2010 to 2017 are Black and Hispanic

From @fig-victim-race-bar and @tbl-victim-race-table it shows that Black and Hispanic victims make up more than 1,000 unsolved cases compared to other victims of other races who have under 1,000 unsolved cases. However, Black victims makes up the majority of unsolved homicide at 2960 cases with only 1103 solved homicides. Based on @tbl-model-result and @fig-model-result-ci, our model shows that there is a slight likelihood that being Black and Hispanic relative to being White impacts a homicide case going unsolved. @citeChavis had found that the Chicago Police Department was under scrutiny for another issue with their gang databases being inaccurate and targeting mostly Black and Latinx residents. @citeChavis also mentions that Chicago is one of the United States' most segregated cities both economically and racially. This suggest that Black and Hispanic victims could be facing systematic biases in homicide case closures and their cases being solved.

@@

In order to reproduce the results in the paper, first run the 00-install_packages.R in the scripts folder located in this paper’s GitHub repository. Then run the other scripts based on the number at the beginning of the script name.

>>>>

The dataset we used for this paper comes from The Washington Post's GitHub repository, "How The Post mapped unsolved murders", which is also known as the "Unsolved Homicide Database" [@datahomicides]. We used the statistical programming language R [@citeR], tidyverse [@citetidyverse], janitor [@citejanitor], lubridate [@citelubridate], dplyr [@citedplyr], ggplot2 [@citeggplot2], arrow [@citearrow], testthat [@citetestthat], and knitr [@citeknitr] to retrieve, clean, test, and analyze the dataset. To construct, test, and analyze our model, we used the following packages: rstanarm [@citerstanarm], modelsummary [@citemodelsummary], and car [@citecar]. The causal model diagram created to understand the relationship between the predictor variables, the outcome variable, and a confounder used DiagrammeR [@citeDiagrammeR], rsvg [@citersvg], DiagrammeRsvg [@citeDiagrammeRsvg] and png [@citepng].

The Washington Post's Unsolved Homicide Database is a dataset compiled by Washington Post reporters that contains over 52000 homicides in the United States (US) from 50 of the largest US cities from 2007 to 2017 [@datahomicides]. This dataset includes information about the victim such as their name, sex, race, and age as well as geographic and temporal information of the homicide. The Washington Post was interested in using the information compiled to map unsolved homicides across the United States in major cities from 2007 to 2017 [@citewashingtonpostmap]. Another dataset we had considered using was another one compiled by The Washington Post on school shootings across the US and approaching our problem with a different perspective on the characteristics of the perpetrator [@citeSchoolShootingsWashington]. However, due to there being only 416 observations with numerous observations missing information, we forgo using the dataset.

We retrieved the raw dataset from The Washington Post using a script that downloads the CSV file from their GitHub repository. The raw dataset contains 52,179 observations with each observation being a homicide case. However, since we narrowed our scope to focus on one of the two most populated US cities, Los Angeles and Chicago, the number of observations in our cleaned dataset ended up being 6,307. Our data look as follows:

@@

@tbl-dataset-preview and @tbl-dataset-statistics indicate our variables of interest, which are the following: victim_race, victim_age, victim_sex, city, disposition, year, month, and arrest_was_not_made. victim_race represents the race of the homicide victim, which can be "White" (376 observations), "Hispanic" (1,763 observations), "Black" (4,063 observations), "Asian" (43 observations), and "Other" (62 observations). victim_age signifies the age of the homicide victim at the time of their death and is defined as a whole number. @tbl-dataset-statistics shows that the mean victim_sex is the sex of the homicide victim where they are identified as a "female" or "male". The city variable defines the city the victim is reported to have been found in. The disposition variable is the specific status of a homicide case where a case can fall in either of the following three statuses: "Closed by arrest", "Open/No arrest", and "Closed without arrest". The year variable represents a year from 2010 to 2017 that indicates the year the homicide took place. Following this, the month variable represents the month a homicide took place. arrest_was_not_made is a variable that was constructed based on the disposition variable with "Closed by arrest" being converted to a 0 and "Open/No arrest" and "Closed without arrest" being converted to a 1. arrest_was_not_made indicates the status of a homicide case as either being unsolved, which is denoted by a 1, and solved, which is denoted by a 0.

However, our dataset has limitations. There was only data available from 2010 onwards for Los Angeles provided by The Washington Post and that limited the number of homicides we could look at for both Chicago and Los Angeles. Also, not all victims were able to be identified in some homicide cases and unknown attributes of the victim such as their age, sex, and gender were indicated with the text "Unknown" in the dataset. However, this causes issues with data type compatibility such as the value "Unknown" being a character type being under the victim_age column, which has a data type of integer. This would lead to issues with our model providing accurate estimates. We decided to remove cases during our data cleaning where victim's demographic information is missing at least one of the three columns, victim_age, victim_sex, victim_race.

@@

The Federal Bureau of Investigation (FBI) has a program called the Uniform Crime Reporting (UCR) Program to generate statistics for the public [@citeFBIinfo]. The FBI originally had a system under the UCR called Summary Reporting System (SRS), which obtained details about different crimes taking place from law enforcement agencies nationwide such as victim information [@citeFBIinfo]. However, the SRS was replaced with a new system called the National Incident-Based Reporting System (NIBR) in 2021 that obtained more details about various crimes [@citeFBIinfo]. For The Los Angeles Police Department, after a homicide case occurs, the investigating team handwrite information into physical crime reports with details like the type of crime, the premise the crime occurred at, and the age and ethnicity of the victim [@citeLApolice]. The crime reports are then transcribed into a digital format and then sent to the SRS monthly [@citeLApolice]. After a homicide occurs in Chicago, The Chicago Police Department use a system called the Chicago Police Department's CLEAR (Citizen Law Enforcement Analysis and Reporting) system to report on details such as the victim, if an arrest was made or not, and the location the crime took place at [@citeChicagoPolice]. This information is then reported to the FBI monthly under the UCR program [@citeChicagoPolice].

Reporters from the Washington Post then obtained the data from the FBI specifically on homicides from the 50 largest US cities from 2007 to 2017, which can be accessed through the UCR publications page on the FBI website [@datahomicides]. They selected the 50 largest cities based the size of the city in 2012 [@datahomicides]. They also obtain data about homicide counts and closure rates through papers and compare these values with the ones from the FBI dataset for accuracy [@datahomicides]. The Washington Post also used public records like medical examiner reports, death certificates, and court records, to fill in any missing information since some departments only report partial information to the FBI [@datahomicides]. The Washington Post defined cases to be closed without arrest when they are reported by the police as "exceptionally cleared" where there is evidence of who the perpetrator is but an arrest is not possible [@datahomicides]. They also define cases to be closed by arrest if the police reported it to be and other cases are defined to be open/no arrest [@datahomicides].

## Outcome variable
The outcome variable we are interested in looking at with our model and our analysis is the arrest_was_not_made variable. We use this variable to compare homicide case characteristics of solved and unsolved homicides from 2010 to 2017.

@@

# Model {#sec-model}
The model we implemented was a Bayesian logistic regression model. This model was constructed after we saw patterns between homicide case information and a homicide case going unsolved in our analysis. We are interested in seeing if certain characteristics of a homicide case such as the victim's perceived characteristics (sex, gender, age) and the year and city the victim is found impact the likelihood of their case going unsolved.

@@

We define $y_i$ to be the status of the homicide case where 1 means the homicide is unsolved (case is still open / been closed without arrest) and 0 means the homicide is solved (case has been closed with arrest). $\pi_i$ represents the probability of the homicide being solved. $\mbox{logit}(\pi_i)$ indicates the log odds of the homicide being unsolved. Now looking at the coefficients $b_i$ and predictor variables, $\beta_0$ is the intercept of our model and is the log odds when all predictor variables are equal to 0. $\mbox{victim\_race}_i$ signifies the race of the victim, which could be either "Asian", "Black", "Hispanic", "White", and "Other". In our model, we use "White" as the baseline for $\mbox{victim\_race}_i$ to see if being part of a minority impacts if the case goes unsolved or not. $\beta_1$ is the coefficient that represents the log odds when $\mbox{victim\_race}_i$ changes. $\mbox{victim\_age}_i$ is the victim's age, which is a whole number. $\beta_2$ is the log odds when $\mbox{victim\_age}_i$ increases by 1 year. $\mbox{victim\_sex}_i$ represents the sex of the victim, where in the dataset it is either "female" or "male" and $\beta_3$ is the log odds when $\mbox{victim\_sex}_i$ changes. $\mbox{city}_i$ is the city the victim was reported to be killed in, which from our dataset could be either "Chicago" or "Los Angeles". $\beta_4$ is the coefficient that stands for the log odds when $\mbox{city}_i$ changes. We define $\mbox{year}_i$ to be the year the homicide was reported to have occured from 2010 to 2017. $\beta_5$ indicates the log odds when $\mbox{year}_i$ increases by 1 year.

Our model runs in R [@citeR] using the rstanarm package [@citerstanarm]. For our model's priors, we use the default priors provided by rstanarm [@citerstanarm]. Diagnostics for the model such as a posterior predictive check, trace and Rhat plots can be found in @sec-model-details.

@@

In our model, we assumed there is a relationship between the outcome variable, homicide is unsolved, with homicide case information like demographic (sex, race, and age of victim), geographic (city), and temporal (year) information, which are the predictor variables. For the demographic data, we decided to keep the grouping provided by The Washington Post such as "female" and "male" for the victim's sex and "White", "Black", "Hispanic", "Asian", and "Other for the victim's race. However, we did remove victims who have any demographic information that fall under the "Unknown" grouping from our dataset and did not run our regression model on these observations. The reason for this is due to factors such as the data type of our predictors and keeping consistency across all observations by removing any unknown values. For example, our predictor variable, victim's age is a integer data type but it contained the string "Unknown" in the raw dataset. So the values containing "Unknown" are removed from our final dataset and not used to train our model. Since our outcome variable is binary for our model, we constructed the arrest_was_not_made column for it based on the disposition variable to indicate if an arrest was made or not or in order words, the homicide is solved or unsolved. We considered dispositions with values like "Open/No arrest" and "Closed without arrest" to be 1 and "Closed by arrest" to be 0 for the arrest_was_not_made column.

We used a logistic regression model in the Bayesian framework since our outcome is binary and it predicts if a homicide is unsolved or not. Another model we considered is a logistic regression model with an instrumental variable. Introducing a instrumental variable into our model could have potentially provided a more accurate model and given us more consistent coefficient estimates as noted by @citeucdavis. However with the available information we had about each case, there was no candidate instrumental variable that impacted at least one variable in our data and not influence the outcome of the case being solved or unsolved. Thus, the model would fail the "Exclusion Restriction" assumption mentioned by @citetellingstories. We also went through different pairs and groupings of variables and found there was not a single strong relationship between variables that is relevant and consequently failing @citetellingstories's "Relevance" assumption. We also have known treatment variable/predictor variables that can be used to measure the outcome variable and therefore, the instrumental variable is less likely to be necessary [@citetellingstories].

Performing root mean square error (RSME) calculations on our model in @tbl-model-result yields a RSME value of about 0.45. RSME represents the difference between the model's predicted values and observed values with 0 indicating that the predicted and observed values are the same [@citejim]. Since the model's RSME value is close to 0, we can say our model is able to predict values with less error compared to other models that have a higher RSME value.

@@

However, our model has limitations and there are situations where this model would not work. @fig-causal-model shows that there is a confounder, "resources of investigating team" between the predictor variables, city and year and the outcome variable, case being unsolved. We define "resources of the investigating team" to include any of the following: the amount of people on the team investigating the case, time allotted to investigate the case, amount of open cases for the team, cost, skill and education levels of members, etc. We currently do not have any information available about the resources of the investigating team for the case and further investigation is needed. We proposed an idealized survey we would conduct to collect the necessary data to further understand the relationship between homicide case characteristics and unsolved homicides in @sec-ideal-survey. If we have information available about the resources of the investigating team for the case, the current model would not work and would need to be revised. This is due to the interaction between the confounder with the city, year, and outcome variable.

@@

@fig-disposition-bar and @tbl-disposition-table shows that the number of homicide cases that are closed by arrest at 1,109 cases is close to the number of homicide cases that are open at 1,087 cases from 2010 to 2017. On the hand, there is a larger difference for Chicago between the number of homicides that are closed by arrest at 947 cases with the ones that are closed without arrest and open/no arrest at 216 and 2,948 cases, respectively.

@@

@fig-victim-age-histogram shows that the distribution of age of homicide victims is relatively uniform. It appears that the victims of unsolved and solved homicides are about the same age if one were to not consider ages above 75. However, if we were to only consider age 75 onwards, it appears that the distribution is a bit right-skewed. We can see that from @tbl-victim-age-statistics, the youngest homicide victim is 1 years old and the oldest is 94 years old. The median age of homicide victims is 44 years old with the mean age being 45 years old.

@@

```{r}
#| label: fig-model-result-ci
#| fig-cap: The credible intervals (line) for coefficient estimates (dot) of predictor variables for homicides that go unsolved from 2010 to 2017.
#| echo: false
#| warning: false
#| message: false
# Specify credible interval to be 90% and present model results
modelplot(unsolved_model_results, conf_level = 0.9) +
  labs(x = "90 per cent credible interval")
```
@tbl-model-result and @fig-model-result-ci present the results from our logistic regression model in the Bayesian framework and the 90% credible intervals of each predictor, respectively. From @tbl-model-result, the intercept $\beta_0$ of -165.865 indicates that when the homicide victim is White, since "White" is the baseline for the victim_race predictor, and victim_age is 0 and year being 2010 with its coefficient being 0 and the victim's sex is female, the homicide is more likely to be solved. @tbl-model-result indicates that the coefficient estimate of victim_race $\beta_1$ is 0.168 when the victim is Asian. This means that the log odds of a homicide case being unsolved increase by 0.168 when the victim is Asian while other predictor variables stay fixed. Following this, when the victim's race is Black or Hispanic, the likelihood of a homicide being unsolved increases by 0.567 and 0.434, respectively. This indicates that the likelihood of a homicide going unsolved increases if the victim's race is Asian, Black, or Hispanic relative to a victim's race being White. On the other hand, when the victim's race falls under "Other", the log odds of a homicide case being unsolved decreases by 0.296. This indicates that the likelihood of a homicide being unsolved decreases if the victim's race falls under "Other" relative to a victim's race being White. However, since the 90% credible interval for the coefficient estimate of victim_race appears to be close to 0 as seen in @fig-model-result-ci, this means that a victim's race has a weak likelihood of impacting the outcome of their case being unsolved.

@tbl-model-result shows that since the coefficient estimate of victim_age $\beta_2$ is -0.006, this means the log odds of a homicide case being unsolved decreases by 0.006 when the age of the victim increases by 1 year as other predictor variables stay constant. This means victims of unsolved homicides are likely on the younger side. @fig-model-result-ci also indicates that $\beta_2$'s 90% credible interval is close to 0 or includes 0 in its interval, indicating there is a chance the victim's age does not have an influence on the outcome of a homicide case being unsolved. With the coefficient estimate of victim_sex $\beta_3$ being 0.662 in @tbl-model-result, it indicates that the log odds of a homicide case being unsolved increases by 0.662 when the victim is male while other predictors are fixed. This suggests that the likelihood of a homicide case being unsolved increases when the victim is a male. Looking at @fig-model-result-ci, the credible interval for $\beta_3$ is also slightly above 0. @tbl-model-result also shows that the log odds of a homicide being unsolved decreases by 1.089 as indicated by the coefficient estimate $\beta_4$ when the city is set to Los Angeles with other predictors being fixed. This indicates that more unsolved homicide cases are likely to occur in Chicago instead. @fig-model-result-ci also shows that $\beta_4$'s credible interval is below 0. Looking at the year predictor and its coefficient estimate, the log odds of a homicide case being unsolved increases by 0.083 as noted by the coefficient estimate $\beta_5$ when the year increases by 1 year while the other predictors stay constant. This indicates that cases are more likely to be unsolved in the later years between 2010 to 2017. @fig-model-result-ci also shows that $\beta_5$'s credible interval is slightly above 0 or includes 0 indicating that the year has a weak likelihood of indicating if a homicide is unsolved.

@@

In @sec-results, we analyzed homicide case information like the victim's age, sex, and race and the year and city where the homicide case occurred to see if there were any differences between homicide case information between unsolved and solved homicides from the year 2010 to 2017. We also built a Bayesian logistic regression model in @sec-model-results, to see the likelihood of the homicide case information impacting the homicide case outcome of going unsolved.

## Chicago is likely to have more unsolved homicides than Los Angeles with a surge in unsolved cases from 2016 onwards

From @fig-city-bar and @tbl-city-table, it indicates that while Los Angeles has slightly more homicide cases that are solved than unsolved, 77% of Chicago's homicide cases are unsolved from 2010 to 2017. This is also seen in @tbl-model-result, where our model indicates that unsolved homicides in Chicago are likely to go unsolved compared to Los Angeles. This suggests potential issues with how the Chicago Police Department is managing homicide case closure. From @fig-month-bar and @fig-year-bar, we saw that most unsolved homicide cases occur in the middle and later half of the year and that there was a jump in unsolved homicides in 2016 and 2017 compared to past years. @citeCNNChicago reported on January 2, 2017, that 2016 was the year with the most number of homicides in the past 19 years due to gun violence in the city. This suggests that most of the unsolved cases could have occurred due to police departments like the Chicago Police Department being unable to keep up with the surge of homicide cases.

@@

@fig-victim-age-histogram indicates that the age of the homicide victim likely does not impact the outcome of their case be unsolved due to the uniformity of the distribution of ages. This is further seen in @fig-model-result-ci, where there is a chance the victim's age does not influence the outcome of a homicide case in our model. This implies that the age of the homicide victim does not impact if their case is unsolved or not. However, @fig-model-result-ci and @fig-victim-sex-bar indicate that female victims had a higher proportion of their homicide cases being solved while on the other hand, male victims had a higher proportion of their homicide cases going unsolved. This suggesst that males are more likely to have their cases go unsolved.

## Disproportionate number of homicide victims in unsolved cases in Chicago and Los Angeles from 2010 to 2017 are Black and Hispanic

From @fig-victim-race-bar and @tbl-victim-race-table it shows that Black and Hispanic victims make up more than 1,000 unsolved cases compared to other victims of other races who have under 1,000 unsolved cases. However, Black victims make up the majority of unsolved homicides at 2,960 cases with only 1,103 solved homicides. Based on @tbl-model-result and @fig-model-result-ci, our model shows that there is a slight likelihood that being Black and Hispanic relative to being White impacts a homicide case going unsolved. @citeChavis had found that the Chicago Police Department was under scrutiny for another issue with their gang databases being inaccurate and targeting mostly Black and Latinx residents. @citeChavis also mentions that Chicago is one of the United States' most segregated cities both economically and racially. This suggests that Black and Hispanic victims could be facing systematic biases in homicide case closures and their cases being solved.

@@

In order to reproduce the results in the paper, first run the 00-install_packages.R in the scripts folder located in this paper’s GitHub repository. Then run the other scripts based on the number at the beginning of the script's name.
